---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# The examples run in a folder of their own, so the printed output carries
# short relative paths and stays the same on any machine.
readme_dir <- tempfile("obt-readme")
dir.create(readme_dir)
knitr::opts_knit$set(root.dir = readme_dir)
```

# obt

<!-- badges: start -->
[![R-CMD-check](https://github.com/epicentre-msf/obt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epicentre-msf/obt/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`obt` drives the OutbreakTools designer and the linelist workbook from R. The
steps normally done by clicking through Excel — download a release, fill a
setup, add a geobase, generate a linelist, run an export — are written as R
code and run in one go.

Each verb records one operation takes the `obt` object as input. 
The verbs chain with `|>`. Nothing reaches
the disk while a recipe is built. `obt_commit()` is the one function that
opens Excel.

## Installation

```r
# install.packages("pak")
pak::pak("epicentre-msf/obt")
```

`obt` reads closed `.xlsb` workbooks through
[readxlsb](https://github.com/velofrog/readxlsb), which comes from GitHub.
`pak` installs it from the `Remotes:` field.

Opening Excel needs **macOS or Windows**, with Microsoft Excel installed. On
any other system the package loads, recipes build and print, and
`obt_commit()` is turned down with a warning message.

On macOS the package copies a run into Excel's own folder, because Excel for
Mac opens a file there without asking permission. Reaching that folder needs Full Disk
Access for the app running your R session. If it can not be reached, the run
works in your own working folder and Excel asks permission access for each file/folder it opens.

`obt_platform()` reads all of the run and prints what the machine can do: the
system, whether a recipe can run, the staging folder, the version of
Excel installed, and the app your R session runs in.

```{r platform, eval = FALSE}
obt_platform()
```

## Download the release

Name a working folder, choose the channel, and the run downloads the zip into
that folder and unzips it. `"main"` is the stable release and `"dev"` is the
development build.

```{r goal-one}
library(obt)

run <- obt(folder = "measles-2026") |>
  obt_designer_add(type = "main")

run
```

Execute the run with `obt_commit()`
```{r goal-one-commit, eval = FALSE}
obt_commit(run)
```

Everything the run downloads or writes lives under the one folder, in a fixed
layout.

```{r paths}
names(obt_paths(run))
```

```{r inputs, include = FALSE}
file.copy(
  system.file("extdata", "generic-test-setup.xlsb", package = "obt"),
  "setup.xlsb"
)
invisible(file.create("geobase.xlsx"))
invisible(file.create("measles-2026.xlsb"))
invisible(file.create("cases.xlsx"))
```

## Turn a setup into a linelist

Fill the release's empty setup from a setup you already have, add a geobase,
choose the two languages, and generate.

```{r goal-two}
run <- obt(folder = "measles-2026") |>
  obt_designer_add(type = "main") |>
  obt_setup_convert(from = "setup.xlsb") |>
  obt_designer_geobase(path = "geobase.xlsx") |>
  obt_designer_languages(dict = "English", form = "ENG") |>
  obt_designer_generate(name = "measles-2026")
```

`obt_describe()` shows every value the run will use and every operation with
its arguments. Read it before `obt_commit()`.

```{r describe}
obt_describe(run)
```

```{r goal-two-commit, eval = FALSE}
finished <- obt_commit(run)
```

A verb called by mistake comes back out. `obt_remove()` takes one step, named
by the number or the type the description prints beside it, and `obt_pop()`
takes the last one off.

```{r remove}
obt_remove(run, step = "designer-geobase") |>
  obt_operations()
```

`obt_commit()` runs the operations in the order the verbs recorded them,
prints each one as it starts and ends, stops at the first failure naming the
operation and the reason. Every run writes one text log under `log/`.

Your own setup and geobase are inputs. The package copies them into the
working folder and works on the copies.

## Working on one workbook

Two verbs narrow a recipe to a single workbook, and each holds the operations
that run inside it. `obt_setup()` holds a setup; `obt_linelist()` holds a
linelist that is already built. Both take the file itself, or an `obt` recipe
to narrow.

```{r linelist}
obt_linelist(from = "measles-2026.xlsb", folder = "measles-2026") |>
  obt_linelist_geobase(path = "geobase.xlsx") |>
  obt_linelist_import(from = "cases.xlsx") |>
  obt_describe()
```

A linelist verb takes a linelist recipe and a designer verb takes an `obt`
one, and a setup verb a setup recipe.

## The rest of the surface

| Area | Verbs |
| --- | --- |
| The recipe | `obt()`, `obt_paths()`, `obt_commit()` |
| Reading a recipe | `obt_summary()`, `obt_operations()`, `obt_describe()` |
| Correcting a recipe | `obt_remove()`, `obt_pop()` |
| Adding the obt files | `obt_designer_add()` |
| Setup operations | `obt_setup()`, `obt_setup_convert()`, `obt_setup_export()`, `obt_setup_import()`, `obt_setup_tags()` |
| Fake records | `obt_fake()` |
| Languages | `obt_setup_languages()`, `obt_designer_languages()`, `obt_languages()` |
| Designer operations | `obt_designer_geobase()`, `obt_designer_generate()` |
| Linelist operations | `obt_linelist()`, `obt_linelist_geobase()`, `obt_linelist_import()`, `obt_linelist_export()` |
| The platform of the run | `obt_platform()`, `obt_staging_clean()` |
| Message boxes in files | `obt_silent_get()`, `obt_silent_set()` |


## Licence

MIT. See `LICENSE.md`.
