---
title: "Build a recipe, then run it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a recipe, then run it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The examples run in a folder of their own, so the printed output carries
# short relative paths and stays the same on any machine.
vignette_dir <- tempfile("obt-vignette")
dir.create(vignette_dir)
knitr::opts_knit$set(root.dir = vignette_dir)
```

```{r setup}
library(obt)
```

The `obt` package provides an automated interface to most of the input/output operations done with [outbreak-tools](https://github.com/epicentre-msf/outbreak-tools) files (setup, designer, linelist).

First you build a **recipe**: a working folder and a list of operations, held
in an ordered R object. Then you read it, correct it, and hand it to
`obt_commit()`, which is the one function that opens Excel and execute them.

Everything in this vignette runs with no Excel and writes nothing to disk.

## The recipe

`obt()` starts one recipe. It takes the working folder and wait for a list
of instructions

```{r start}
run <- obt(folder = "measles-2026")
run
```

The folder is recorded here and is created/read when the run starts, 
so a recipe can be built on a machine where the folder does not exist yet. 
`obt_paths()` shows the layout the run will use.

```{r paths}
paths <- obt_paths(run)
names(paths)
```

Each verb adds one operation to the recipe. They can chain with the pipe.

## The OutbreakTools files

`obt_designer_add()` records the OutbreakTools folder to use for the operations. There are two main options: `"main"` is the stable
folder and `"dev"` is the development  one. The folder is a zip with the setup, the designer and other supporting files. The zip is downloaded in the `obt/` at run time, in the working folder.

```{r designer}
run <- run |>
  obt_designer_add(type = "main")

run
```

## A setup, a geobase and a linelist

`obt_setup_convert()` takes a setup workbook you already have and fills a
fresh copy of the empty setup the release ships. Your own file is an input:
the run copies it in and works on the copy.

```{r inputs, include = FALSE}
file.copy(
  system.file("extdata", "generic-test-setup.xlsb", package = "obt"),
  "setup.xlsb"
)
```

```{r convert}
run <- run |>
  obt_setup_convert(from = "setup.xlsb")
```

A geobase is optional. `obt_designer_geobase()` points the run to the path of a geobase.

```{r geobase}
invisible(file.create("geobase.xlsx"))

run <- run |>
  obt_designer_geobase(path = "geobase.xlsx")
```

Two languages go into a linelist. `dict` is
the dictionary language, a column of the setup's translation table. `form` is
the interface language: the buttons, the messages and the sheet names.

```{r languages}
run <- run |>
  obt_designer_languages(dict = "English", form = "ENG")
```

`obt_setup_languages()` reads the dictionary languages a setup file holds.

```{r setup-languages}
obt_setup_languages("setup.xlsb")
```

Last, the linelist itself. `name` is the name of the file the designer writes
under `linelist/`. `password` is what the linelist opens with and
`debug_password` what its sheets are protected with. Both are left out here,
so the linelist opens with none.

```{r generate}
run <- run |>
  obt_designer_generate(name = "measles-2026")
```

## Read the recipe before it runs

Three functions read a recipe.

`obt_summary()` is the one line: the folder, the count, and whether every
operation can run. An operation written before the workbooks carry its entry
point is counted here as waiting instead; nothing waits today.

```{r summary}
obt_summary(run)
```

`obt_operations()` says what each step will do.

```{r operations}
obt_operations(run)
```

`obt_describe()` gives more details: every value the run will use, then every
operation with its arguments. This is what to look at before Excel opens.

```{r describe}
obt_describe(run)
```

## Correct a recipe

You can remove operations with `obt_remove()`. It takes one step, named
by the step number or a character which is the type of the operation. `obt_pop()` takes
the last operation off.

```{r remove}
run |>
  obt_remove(step = "designer-geobase") |>
  obt_pop() |>
  obt_operations()
```


## Work on one workbook

Some operations run inside a single workbook, and each of those workbooks has
a recipe class of its own. A class holds the verbs that touch that workbook
and nothing else, so a recipe says which workbook it drives before Excel
opens.

`obt_setup()` holds a setup. Its three verbs are `obt_setup_export()`, which
writes the setup out as an `.xlsx`, `obt_setup_import()`, which reads one back
in, and `obt_setup_tags()`, which updates the translation tags.

```{r setup-recipe}
obt_setup(from = "setup.xlsb", folder = "measles-2026") |>
  obt_setup_tags() |>
  obt_setup_export()
```


`obt_linelist()` holds a linelist that is already built. Its verbs are
`obt_linelist_geobase()`, which reads a geobase in, `obt_linelist_import()`,
which reads a file exported by linelist, and `obt_linelist_export()`, which
writes one out.

```{r linelist-inputs, include = FALSE}
invisible(file.create("measles-2026.xlsb"))
invisible(file.create("cases.xlsx"))
```

```{r linelist-recipe}
obt_linelist(from = "measles-2026.xlsb", folder = "measles-2026") |>
  obt_linelist_geobase(path = "geobase.xlsx") |>
  obt_linelist_import(from = "cases.xlsx") |>
  obt_describe()
```

A linelist built with a password is named with `from`, and `password` is what
every verb of the recipe opens it with. `obt_silent_get()` and
`obt_silent_set()` take the same `password`.

Both take the workbook itself, as above, or an `obt` recipe to narrow. Narrowing keeps the working folder and the operations of that class, and
leaves the rest behind. A recipe that names no workbook of its own reads the
one file the working folder holds.

```{r narrowing}
obt_linelist(run)
```

## Generate fake datasets

`obt_fake()` reads the setup and create a table of records for every
worksheet it names, written into one `.xlsx`.

It runs on an `obt` recipe or a setup one, and it is the only verb that opens
no workbook: the setup is read and the file is written from R, so it works on
any machine. Give it a `seed` and two runs of the recipe answer the same
records.

```{r fake-recipe}
obt_setup(from = "setup.xlsb", folder = "measles-2026") |>
  obt_fake(n = 200, seed = 1)
```

## Execute the operations

`obt_commit()` runs the operations in the order the verbs recorded them. It
creates the working folder, downloads what the recipe asked for, opens Excel
where an operation needs it, and stops at the first failure with the operation
and the reason. Every run writes one text log under `log/`.

```{r commit, eval = FALSE}
finished <- obt_commit(run)
```

A run can be executed if the platform and the excel versions allow it. `obt_platform()` provides informations about the current run environment: the operating system, and on macOS whether the package can reach Excel's
own folder. It prints the answer for the machine you are on, together with the
version of Excel installed and the app your R session runs in.

```{r platform, eval = FALSE}
obt_platform()
```

Execution works only on macOS or Windows. Everywhere else a recipe still builds,
prints and tests, and `obt_commit()` is the one thing that will not run.